Programming Language Foundations in Agda

### 2019-07-21 papers/scp/PLFA.tex
fix spelling

  What did we learn from writing PLFA? First, that it is possible. One
  might expect that without proof tactics that the proofs become too
  long, but in fact proofs in PLFA are about the same length as those
  in SF. Proofs in Coq require an interactive environment to be
  understood, while proofs in Agda can be read on the page.  Second,
  that constructive proofs of preservation and progress give immediate
  rise to a prototype evaluator. This fact is obvious in retrospect
  but it is not exploited in SF (which instead provides a separate
  normalise tactic) nor can I find it in the literature.  Third, that
  using extrinsicly-typed terms is far less
  perspicuous than using intrinsicly-typed terms. SF uses the former
  presentation, while PLFA presents both; the former uses about 1.6 as
  many lines of Agda code as the latter, roughly the golden ratio.

@@

Section~5 claims that extrinsicly-typed terms should be avoided in favour of
intrisicly-typed terms.  PLFA develops lambda calculus with both,
permitting a comparison.  It turns out the
former is less powerful---it supports substitution only for closed
terms---but significantly longer---about 1.6 times as many lines of code,
roughly the golden ratio.

@@

\paragraph{Lambda: Introduction to lambda calculus}
Introduces lambda calculus, using a representation with named
variables and extrinsicly typed. The language used is PCF
\citep{Plotkin-1977}, with variables, lambda abstraction, application,
zero, successor, case over naturals, and fixpoint. Reduction is
call-by-value and restricted to closed terms.

@@

\paragraph{DeBruijn: Intrinsicly-typed de Bruijn representation}
Introduces de Bruijn indices and the intrinsicly-typed representation.
Emphasis is put on the structural similarity between a term and its
corresponding type derivation; in particular, de Bruijn indices
correspond to the judgment that a variable is well-typed under a given
environment.

\paragraph{More: More constructs of simply-typed lambda calculus}
Introduces product, sum, unit, and empty types; and explains lists and let bindings.
Typing and reduction rules are given informally; a few
are then give formally, and the rest are left as exercises for the reader.
The intrinsicly-typed representation is used.

@@

\paragraph{Inference: Bidirectional type inference}
Introduces bidirectional type inference, and applies it to convert from
a representation with named variables and extrinsicly typed
to a representation with de Bruijn indices and intrinsicly typed. Bidirectional
type inference is shown to be both sound and complete.

\paragraph{Untyped: Untyped calculus with full normalisation}
As a variation on earlier themes, discusses an untyped (but intrinsicly
scoped) lambda calculus.  Reduction is call-by-name over open terms,
with full normalisation (including reduction under lambda terms).  Emphasis
is put on the correspondence between the structure of a term and
evidence that it is in normal form.

@@

PLFA and SF differ in several particulars.  PLFA begins with a computationally
complete language, PCF, while SF begins with a minimal language, simply-typed
lambda calculus with booleans.  PLFA does not include type annotations in terms,
and uses bidirectional type inference, while SF has terms with unique types and
uses type checking.  SF also covers a simple imperative language with Hoare
logic, and for lambda calculus covers subtyping, record types, mutable
references, and normalisation---none of which are treated by PLFA.  PLFA covers
an intrinsicly-typed de Bruijn representation, bidirectional type inference,
bisimulation, and an untyped call-by-name language with full
normalisation---none of which are treated by SF.  The new part on
Denotational Semantics also covers material not treated by SF.

@@

% TODO: rewrite use of 'Philip' to reflect all authors?
Philip was therefore surprised to realise that any constructive proof of
progress and preservation \emph{automatically} gives rise to such a
prototype.  The input is a term together with evidence the term is
well-typed.  (In the intrinsicly-typed case, these are the same thing.)
Progress determines whether we are done, or should take another step;
preservation provides evidence that the new term is well-typed, so we
may iterate. In a language with guaranteed termination such as Agda, we cannot
iterate forever, but there are a number of well-known techniques to
address that issue; see, e.g., \citet{Bove-and-Capretta-2001},
\citet{Capretta-2005}, or \citet{McBride-2015}.
We use the simplest, similar to McBride's \emph{petrol-driven} (or
\emph{step-indexed}) semantics: provide a maximum number of steps to
execute; if that number proves insufficient, the evaluator returns the
term it reached, and one can resume execution by providing a new
number.

@@

% TODO: consider rewriting this?
The second part of PLFA first discusses two different approaches to
modeling simply-typed lambda calculus.  It first presents
terms with named variables and and extrinsic typing relation and
then shifts to terms with de Bruijn indices that are intrinsicly
typed.  The names \emph{extrinsic} and \emph{intrinsic} for these
two approaches are taken from \citet{Reynolds-2003}.
Before writing the text, Philip had thought the two approaches
complementary, with no clear winner.  Now he is convinced that the
intrinsicly-typed approach is superior.

@@

For inventing ideas on which PLFA is based, and for hand-holding, many thanks to
Conor McBride, James McKinna, Ulf Norell, and Andreas Abel.  For showing how
much more compact it is to use intrinsicly-typed terms, thanks to David Darais.  For
inspiring our work by writing SF, thanks to Benjamin Pierce and his coauthors.
For comments on a draft of this paper, an extra thank you to James McKinna, Ulf
Norell, Andreas Abel, and Benjamin Pierce. This research was supported by EPSRC
Programme Grant EP/K034413/1.

>>>>

  What did we learn from writing PLFA? First, that it is possible. One
  might expect that without proof tactics that the proofs become too
  long, but in fact proofs in PLFA are about the same length as those
  in SF. Proofs in Coq require an interactive environment to be
  understood, while proofs in Agda can be read on the page.  Second,
  that constructive proofs of preservation and progress give immediate
  rise to a prototype evaluator. This fact is obvious in retrospect
  but it is not exploited in SF (which instead provides a separate
  normalise tactic) nor can I find it in the literature.  Third, that
  using extrinsically-typed terms is far less
  perspicuous than using intrinsically-typed terms. SF uses the former
  presentation, while PLFA presents both; the former uses about 1.6 as
  many lines of Agda code as the latter, roughly the golden ratio.

@@

Section~5 claims that extrinsically-typed terms should be avoided in favour of
intrisicly-typed terms.  PLFA develops lambda calculus with both,
permitting a comparison.  It turns out the
former is less powerful---it supports substitution only for closed
terms---but significantly longer---about 1.6 times as many lines of code,
roughly the golden ratio.

@@

\paragraph{Lambda: Introduction to lambda calculus}
Introduces lambda calculus, using a representation with named
variables and extrinsically typed. The language used is PCF
\citep{Plotkin-1977}, with variables, lambda abstraction, application,
zero, successor, case over naturals, and fixpoint. Reduction is
call-by-value and restricted to closed terms.

@@

\paragraph{DeBruijn: Intrinsically-typed de Bruijn representation}
Introduces de Bruijn indices and the intrinsically-typed representation.
Emphasis is put on the structural similarity between a term and its
corresponding type derivation; in particular, de Bruijn indices
correspond to the judgment that a variable is well-typed under a given
environment.

\paragraph{More: More constructs of simply-typed lambda calculus}
Introduces product, sum, unit, and empty types; and explains lists and let bindings.
Typing and reduction rules are given informally; a few
are then give formally, and the rest are left as exercises for the reader.
The intrinsically-typed representation is used.

@@

\paragraph{Inference: Bidirectional type inference}
Introduces bidirectional type inference, and applies it to convert from
a representation with named variables and extrinsically typed
to a representation with de Bruijn indices and intrinsically typed. Bidirectional
type inference is shown to be both sound and complete.

\paragraph{Untyped: Untyped calculus with full normalisation}
As a variation on earlier themes, discusses an untyped (but intrinsically
scoped) lambda calculus.  Reduction is call-by-name over open terms,
with full normalisation (including reduction under lambda terms).  Emphasis
is put on the correspondence between the structure of a term and
evidence that it is in normal form.

@@

PLFA and SF differ in several particulars.  PLFA begins with a computationally
complete language, PCF, while SF begins with a minimal language, simply-typed
lambda calculus with booleans.  PLFA does not include type annotations in terms,
and uses bidirectional type inference, while SF has terms with unique types and
uses type checking.  SF also covers a simple imperative language with Hoare
logic, and for lambda calculus covers subtyping, record types, mutable
references, and normalisation---none of which are treated by PLFA.  PLFA covers
an intrinsically-typed de Bruijn representation, bidirectional type inference,
bisimulation, and an untyped call-by-name language with full
normalisation---none of which are treated by SF.  The new part on
Denotational Semantics also covers material not treated by SF.

@@

% TODO: rewrite use of 'Philip' to reflect all authors?
Philip was therefore surprised to realise that any constructive proof of
progress and preservation \emph{automatically} gives rise to such a
prototype.  The input is a term together with evidence the term is
well-typed.  (In the intrinsically-typed case, these are the same thing.)
Progress determines whether we are done, or should take another step;
preservation provides evidence that the new term is well-typed, so we
may iterate. In a language with guaranteed termination such as Agda, we cannot
iterate forever, but there are a number of well-known techniques to
address that issue; see, e.g., \citet{Bove-and-Capretta-2001},
\citet{Capretta-2005}, or \citet{McBride-2015}.
We use the simplest, similar to McBride's \emph{petrol-driven} (or
\emph{step-indexed}) semantics: provide a maximum number of steps to
execute; if that number proves insufficient, the evaluator returns the
term it reached, and one can resume execution by providing a new
number.

@@

% TODO: consider rewriting this?
The second part of PLFA first discusses two different approaches to
modeling simply-typed lambda calculus.  It first presents
terms with named variables and and extrinsic typing relation and
then shifts to terms with de Bruijn indices that are intrinsically
typed.  The names \emph{extrinsic} and \emph{intrinsic} for these
two approaches are taken from \citet{Reynolds-2003}.
Before writing the text, Philip had thought the two approaches
complementary, with no clear winner.  Now he is convinced that the
intrinsically-typed approach is superior.

@@

For inventing ideas on which PLFA is based, and for hand-holding, many thanks to
Conor McBride, James McKinna, Ulf Norell, and Andreas Abel.  For showing how
much more compact it is to use intrinsically-typed terms, thanks to David Darais.  For
inspiring our work by writing SF, thanks to Benjamin Pierce and his coauthors.
For comments on a draft of this paper, an extra thank you to James McKinna, Ulf
Norell, Andreas Abel, and Benjamin Pierce. This research was supported by EPSRC
Programme Grant EP/K034413/1.
